When Risk Hits the Road: Does Danger Drive a Return?

### 2024-12-03 paper/paper.qmd
changes on the color of the plots in variable sections

---
title: "When Risk Hits the Road: Does Danger Drive a Return?"
subtitle: "Motor Theft and Collisions in Toronto Are Driven by [Neighborhood Clustering] and [Environmental Conditions]"
author: 
  - Yingke He
thanks: "Code and data are available at: [https://github.com/ohyykk/Toronto_Motor_Viehicle/tree/main](https://github.com/ohyykk/Toronto_Motor_Viehicle/tree/main)."
date: today
date-format: long
abstract: "This study examines motor vehicle theft and traffic collisions in Toronto, focusing on spatial and temporal patterns,and environmental factors. A composite risk score model is used and reveals that collisions occur more frequently under unfavorable road conditions, such as poor lighting and wet or icy surfaces, and in areas with inadequate traffic control measures. These findings emphasize the need for targeted interventions, including enhanced road infrastructure and lighting in collision-prone zones. Such measures aim to improve safety and security for motorbike riders and owners, contributing to safer urban mobility in Toronto."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

The **Hour of the Day** variable is used as a predictor to analyze temporal trends in motorbike theft occurrences. This variable helps identify whether certain hours are associated with elevated or reduced theft risks. Contributing factors may include decreased monitoring during nighttime hours, increased activity in high-risk areas during specific times, or patterns related to commuter schedules. @plt-hour_summary outlines the distribution of theft incidents across different hours, offering insights into periods with disproportionately high or low occurrences, which could impact the understanding of temporal theft risk dynamics.

```{r}
#| label: plt-hour_summary
#| fig-cap: "Theft Incidents by Hour of the Day"
#| echo: false
#| warning: false
#| message: false
# Load necessary libraries
library(tidyverse)

@@

# Create the plot
ggplot(hour_summary, aes(x = hour, y = theft_count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  scale_x_continuous(breaks = 0:23, labels = sprintf("%02d", 0:23)) + # Ensure 24-hour format
  theme_minimal() +
  labs(

@@

The **Day of the Week** variable is included as a predictor to examine weekly patterns in motorbike theft incidents. This variable helps determine whether theft risks vary across different days of the week. Factors such as increased activity on weekends, reduced monitoring during certain weekdays, or patterns linked to routine schedules may influence these variations. @plot-day_summary summarizes the distribution of theft incidents by day of the week, highlighting any days with notably high or low occurrences.

```{r}
#| label: plot-day_summary
#| fig-cap: "Theft Incidents by Day of the Week"
#| echo: false
#| warning: false
#| message: false

@@

# Create the plot with a different color
ggplot(day_summary, aes(x = day_of_week, y = theft_count)) +
  geom_bar(stat = "identity", fill = "tomato") +
  theme_minimal() +
  labs(

@@

```
#### Premises Type

The **Premises Type** variable serves as a predictor to analyze the contextual settings where motorbike theft incidents occur. This variable helps identify whether certain types of premises, such as residential areas, commercial establishments, or public parking lots, are associated with higher or lower theft risks. Factors like the availability of surveillance, the density of parked motorbikes, or accessibility to offenders may contribute to variations across premises types. @plt-premises_summary provides a breakdown of the number of theft incidents by premises type, highlighting locations with elevated or reduced risks.

```{r}
#| label: plt-premises_summary
#| fig-cap: "Theft Incidents by Day of the Week"
#| echo: false
#| warning: false
#| message: false

@@

# Create the plot for premises type
ggplot(premises_summary, aes(x = reorder(premises_type, -theft_count), y = theft_count)) +
  geom_bar(stat = "identity", fill = "dodgerblue") +
  theme_minimal() +
  labs(
    title = "Theft Incidents by Premises Type",
    x = "Premises Type",
    y = "Number of Theft Incidents"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )

@@

The **Neighborhood ID** variable is a categorical predictor that represents unique identifiers for neighborhoods in the City of Toronto. This variable captures spatial variations in motorbike theft and collision risks, enabling the analysis to identify neighborhood-specific patterns and trends. @tbl-neighborhood_summary provides a list of example Neighborhood IDs, illustrating how this variable is used to represent different areas in the dataset. This summary helps contextualize the role of neighborhoods in understanding collision risk distributions.

@@

The Road Surface Condition variable is a categorical predictor that describes the state of the road at the time of an incident, including categories such as dry, wet, loose snow, slush, ice, packed snow, gravel, spilled liquid, and others. This variable helps analyze how varying surface conditions influence motorbike collision risks. @plot-road_surface_summary illustrates the number of incidents across all these road surface conditions, revealing that the majority of incidents occur on dry roads, followed by wet roads. Other surface conditions, such as snow, ice, or gravel, account for a much smaller proportion of incidents. This distribution suggests that while dry conditions are most common for collisions, likely due to higher traffic volumes, wet and less stable surfaces may pose proportionally greater risks. These insights emphasize the importance of considering all road conditions when designing targeted road safety measures, particularly during adverse weather, to mitigate collision risks effectively.

```{r}
#| label: plot-road_surface_summary
#| fig-cap: "Incidents by Road Surface Condition"
#| echo: false
#| warning: false
#| message: false

@@

# Create the bar plot for all road surface conditions
ggplot(road_surface_summary, aes(x = reorder(road_conditions, -incident_count), y = incident_count)) +
  geom_bar(stat = "identity", fill = "seagreen") +
  theme_minimal() +
  labs(
    title = "Incidents by Road Surface Condition",
    x = "Road Surface Condition",
    y = "Number of Incidents"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )

@@

The Lighting Condition variable is a categorical predictor that represents the type of lighting at the time of an incident, including conditions such as daylight, darkness, artificial lighting, or dawn. This variable is crucial for assessing how visibility levels and lighting environments influence the likelihood of motorbike thefts and collisions. The bar plot @plt-lighting_summary shows that the majority of incidents occur during daylight conditions, which likely reflects higher traffic volumes during daytime hours. Incidents under "Dark" and "Dark, artificial" lighting conditions are significantly fewer but still notable, emphasizing the increased risks associated with reduced visibility. Other conditions, such as "Dawn" and "Dusk," contribute minimally to the total incidents. These insights highlight the importance of targeted safety interventions, such as improved artificial lighting and driver awareness during nighttime and transitional lighting conditions, to mitigate collision risks.
```{r}
#| label: plt-lighting_summary
#| fig-cap: "Incidents by Lighting Conditionk"
#| echo: false
#| warning: false
#| message: false

# Combine "NA" with "Dawn, artificial" and summarize
lighting_summary <- cleaned_traffic_data %>%
  mutate(
    lighting_conditions = case_when(
      lighting_conditions %in% c("NA", "Dawn, artificial") ~ "Dawn, artificial",
      TRUE ~ lighting_conditions
    )
  ) %>%
  filter(lighting_conditions != "Other") %>%  # Exclude "Other"
  group_by(lighting_conditions) %>%
  summarize(incident_count = n()) %>%
  arrange(desc(incident_count))  # Sort by the number of incidents

# Create the updated bar plot
ggplot(lighting_summary, aes(x = reorder(lighting_conditions, -incident_count), y = incident_count)) +
  geom_bar(stat = "identity", fill = "darkorange") +
  theme_minimal() +
  labs(
    x = "Lighting Condition",
    y = "Number of Incidents"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )

@@

The Traffic Control variable categorizes the type of traffic management present at the location of each incident, such as "No Control," "Traffic Signal," or "Stop Sign." This variable enables an analysis of how different traffic control measures correlate with collision risks. As shown in @tbl-traffic_control_summary, the majority of incidents occur in areas with "No Control" (9,021 incidents), followed closely by areas with "Traffic Signal" (8,035 incidents). This suggests that the absence of traffic control mechanisms, as well as high traffic volumes at signalized intersections, may contribute significantly to collision probabilities. Less frequent traffic control types, such as "Stop Signs" (1,464 incidents) or "Pedestrian Crossovers" (208 incidents), are associated with considerably fewer incidents. These findings underscore the importance of implementing and optimizing traffic control measures, particularly in areas without existing controls, to reduce collision risks effectively.
```{r}
#| label: tbl-traffic_control_summary
#| fig-cap: "Summary of Traffic Control"
#| echo: false
#| warning: false
#| message: false

@@

The **Time of Day** variable captures the specific hour and minute when a collision occurred, represented in a 24-hour format. For instance, "2210" corresponds to 10:10 PM, while "315" indicates 3:15 AM. This variable enables a detailed analysis of temporal patterns in collision risks, helping to identify whether certain times of the day are associated with higher or lower probabilities of incidents.

Contributing factors may include reduced visibility during nighttime hours, increased traffic density during peak commuting periods, or varying driver behaviors at different times. @tbl-time_of_day_summary provides examples of time data, offering a comprehensive view of how collision occurrences are distributed throughout the day. Understanding these temporal trends can inform targeted interventions to enhance road safety during high-risk periods.

```{r}
#| label: tbl-time_of_day_summary
#| fig-cap: "5 Examples of Different Times of Day"
#| echo: false
#| warning: false
#| message: false

@@

# Select 5 random examples of time_of_day
time_of_day_examples <- cleaned_traffic_data %>%
  select(time_of_day) %>%
  distinct() %>%               # Ensure unique times
  sample_n(5)                  # Randomly select 5 examples

# Create the table with a frame
time_of_day_examples %>%
  kable(
    col.names = c("Time of Day"),
    align = "c"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "bordered"),
    full_width = FALSE
  )

#### Visibility
The Visibility variable is a categorical predictor that describes the environmental visibility conditions at the time of a collision, such as clear, foggy, or reduced due to rain or snow. This variable is critical for assessing how different visibility levels impact the likelihood of collisions. Poor visibility conditions, such as heavy fog, snow, or rain, can obscure drivers’ ability to perceive road hazards or other vehicles, increasing the probability of incidents. Conversely, clear visibility conditions often correlate with lower collision risks.

@plot-visibility_summary provides a barplot summarizing the number of collision incidents under each visibility condition, illustrating how varying environmental factors contribute to changes in collision probabilities. The plot shows that the majority of incidents occur under clear conditions, with a significantly higher incident count than other conditions. Rain, the second most frequent condition, accounts for a smaller but notable proportion of incidents, while other conditions like snow, fog, and strong winds contribute minimally. These findings emphasize the need for proactive measures such as improving road drainage and visibility aids during adverse weather to mitigate collision risks effectively.

```{r}
#| label: plot-visibility_summary
#| fig-cap: "Incidents by Visibility Condition"
#| echo: false
#| warning: false
#| message: false

@@

# Create the updated bar plot for visibility conditions
ggplot(visibility_summary, aes(x = reorder(visibility_conditions, -incident_count), y = incident_count)) +
  geom_bar(stat = "identity", fill = "cornflowerblue") +
  theme_minimal() +
  labs(
    x = "Visibility Condition",
    y = "Number of Incidents"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )

@@

```{r}
#| label: fig-risk-index-timeseries-year
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "Risk Index Variations in 2020 during Covid-19"
#| fig-pos: H
#| fig-width: 8.5
#| fig-height: 3.5

# Filter data for a specific year (e.g., 2020)
risk_trends_2020 <- cleaned_traffic_data %>%
  filter(year(date) == 2020) %>%
  group_by(date) %>%
  summarize(avg_risk_index = mean(risk_index, na.rm = TRUE))

risk_trends_2020 <- risk_trends_2020 %>%
  mutate(month = month(date, label = TRUE))

ggplot(risk_trends_2020, aes(x = date, y = avg_risk_index)) +
  geom_line(color = "#66471d", size = 0.8) +
  facet_wrap(~month, scales = "free_x") +  # Facet by month
  labs(
    x = "Month",
    y = "Average Risk Index"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 12)
  )

@@

To better understand how environmental factors influence the Risk Index, we examine the distribution of the Risk Index under various road surface conditions. This analysis explores whether different surface types—such as dry, wet, icy, and others—affect the level of risk associated with traffic incidents. The following violin plot visually compares the Risk Index across different road surface conditions, providing insight into how these factors may contribute to the overall risk.
```{r}
#| label: fig-risk-distribution-roadcondition
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "Risk Index Distribution by Road Surface Conditions"
#| fig-pos: H
# Load necessary libraries
library(ggplot2)
library(dplyr)

# Combine "Spilled Liquid" with "Wet" and filter the data
violin_data <- cleaned_traffic_data %>%
  mutate(
    road_conditions = case_when(
      road_conditions == "Spilled liquid" ~ "Wet",
      TRUE ~ road_conditions
    )
  ) %>%
  filter(!is.na(road_conditions), !is.na(risk_index))  # Ensure no NA values

# Create the updated violin plot
ggplot(violin_data, aes(x = road_conditions, y = risk_index, fill = road_conditions)) +
  geom_violin(trim = FALSE) +  # Add violin layers
  scale_fill_brewer(palette = "Set3") +  # Use a color palette for better distinction
  theme_minimal() +
  labs(
    x = "Road Surface Condition",
    y = "Risk Index",
    fill = "Road Conditions"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )

@@

```{r}
#| label: fig-risk-index-lighting-conditions
#| fig-cap: "Risk Index by Lighting Conditions"
#| fig-subcap: ["Low Light (Dark & Artificial)", "Dawn and Artificial Light", "Bright Daylight and Artificial Light", "Other Lighting Conditions"]
#| echo: false
#| warning: false
#| message: false
#| layout-ncol: 2
#| layout-nrow: 2

@@

# First Plot: Dark
plot_dark <- ggplot(merged_data %>% filter(lighting_condition_group == "Dark"), aes(x = risk_index, fill = lighting_condition_group)) +
  geom_density(alpha = 0.6) +
  scale_fill_manual(values = c("Dark" = "#54350a")) +
  labs(x = "Risk Index", y = "Density", title = "Risk Index for Dark Conditions") +
  theme_minimal() +
  theme(legend.position = "none")

# Second Plot: Daylight
plot_daylight <- ggplot(merged_data %>% filter(lighting_condition_group == "Daylight"), aes(x = risk_index, fill = lighting_condition_group)) +
  geom_density(alpha = 0.6) +
  scale_fill_manual(values = c("Daylight" = "#FFA07A")) +
  labs(x = "Risk Index", y = "Density", title = "Risk Index for Daylight Conditions") +
  theme_minimal() +
  theme(legend.position = "none")

# Third Plot: Dawn
plot_dawn <- ggplot(merged_data %>% filter(lighting_condition_group == "Dawn"), aes(x = risk_index, fill = lighting_condition_group)) +
  geom_density(alpha = 0.6) +
  scale_fill_manual(values = c("Dawn" = "#87CEFA")) +
  labs(x = "Risk Index", y = "Density", title = "Risk Index for Dawn Conditions") +
  theme_minimal() +
  theme(legend.position = "none")

# Fourth Plot: Artificial Lighting
plot_artificial <- ggplot(merged_data %>% filter(lighting_condition_group == "Artificial Lighting"), aes(x = risk_index, fill = lighting_condition_group)) +
  geom_density(alpha = 0.6) +
  scale_fill_manual(values = c("Artificial Lighting" = "#A9A9A9")) +
  labs(x = "Risk Index", y = "Density", title = "Risk Index for Artificial Lighting Conditions") +
  theme_minimal() +
  theme(legend.position = "none")

@@

@fig-risk-index-lighting-conditions displays the distribution of the Risk Index across various lighting conditions. The data is broken down into four main lighting categories: Low Light (Dark & Artificial), Dawn and Artificial Light, Bright Daylight and Artificial Light, and Other Lighting Conditions.

@fig-risk-index-lighting-conditions-1 highlights that collision risks are higher under poor visibility, with a concentration of Risk Index values in the moderate-to-high range. These findings indicate that insufficient natural light, coupled with artificial lighting, may contribute to a heightened likelihood of incidents.

>>>>

---
title: "When Risk Hits the Road: Does Danger Drive a Return?"
subtitle: "Risk Peaks in the Morning on Dark, Snow-Packed Roads and in High Urban Density Areas"
author: 
  - Yingke He
thanks: "Code and data are available at: [https://github.com/ohyykk/Toronto_Motor_Viehicle/tree/main](https://github.com/ohyykk/Toronto_Motor_Viehicle/tree/main)."
date: today
date-format: long
abstract: "This study examines motor vehicle theft and traffic collisions in Toronto, focusing on spatial and temporal patterns,and environmental factors. A composite risk score model is used and reveals that collisions occur more frequently under unfavorable road conditions, such as poor lighting and wet or icy surfaces, and in areas with inadequate traffic control measures. These findings emphasize the need for targeted interventions, including enhanced road infrastructure and lighting in collision-prone zones. Such measures aim to improve safety and security for motorbike riders and owners, contributing to safer urban mobility in Toronto."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

The **Hour of the Day** variable is used as a predictor to analyze temporal trends in motorbike theft occurrences. This variable helps identify whether certain hours are associated with elevated or reduced theft risks. Contributing factors may include decreased monitoring during nighttime hours, increased activity in high-risk areas during specific times, or patterns related to commuter schedules. @fig-hour_summary outlines the distribution of bicycle theft incidents across different hours of the day in a 24-hour format. This temporal variable helps the model capture time-dependent patterns in theft occurrences, identifying periods of heightened risk, such as late morning to early afternoon (10:00 to 15:00), and lower-risk periods, such as early morning hours (02:00 to 06:00).By including this variable, the theft index model utilizes the observed trend of thefts being more frequent during morning and early afternoon hours (7:00 to 13:00) while being significantly lower during the early morning hours (02:00 to 06:00). This temporal variability highlights specific periods of increased theft risk, enabling the model to capture these patterns effectively. Understanding this trend allows for better predictions of theft dynamics and offers valuable insights for designing targeted interventions and optimizing resource allocation based on the time of day.

```{r}
#| label: fig-hour_summary
#| fig-cap: "Theft Incidents by Hour of the Day"
#| echo: false
#| warning: false
#| message: false
# Load necessary libraries
library(tidyverse)

@@

# Create the plot
ggplot(hour_summary, aes(x = hour, y = theft_count)) +
  geom_bar(stat = "identity", fill ="darkgrey") +
  scale_x_continuous(breaks = 0:23, labels = sprintf("%02d", 0:23)) + # Ensure 24-hour format
  theme_minimal() +
  labs(

@@

The **Day of the Week** variable is included as a predictor to analyze weekly trends in motorbike theft incidents. @fig-plot-day_summary visualizes the distribution of thefts across the week, revealing that incidents tend to peak on Monday and Tuesday, with slightly lower occurrences on weekends, particularly Saturday and Sunday. This predictor captures potential variations tied to weekly routines, such as higher theft risks during the start of the workweek when urban activity is higher, or reduced risks over the weekend when monitoring or exposure may change. By incorporating this variable, the model can better account for these temporal patterns, enhancing its ability to predict theft incidents based on the day of the week.

```{r}
#| label: fig-plot-day_summary
#| fig-cap: "Theft Incidents by Day of the Week"
#| echo: false
#| fig-width: 8
#| fig-height: 3.5
#| warning: false
#| message: false

@@

# Create the plot with a different color
ggplot(day_summary, aes(x = day_of_week, y = theft_count)) +
  geom_bar(stat = "identity", fill = "darkgrey") +
  theme_minimal() +
  labs(

@@

#### Premises Type

The **Premises Type** variable is included as a predictor to examine the contextual environments where motorbike theft incidents are most prevalent. This variable categorizes thefts by location type, such as outdoor spaces, residential areas, or commercial properties, to identify settings associated with higher or lower risks. @plt-premises_summary highlights that outdoor locations account for the majority of thefts, followed by residential areas like houses and apartments, while locations such as educational and transit premises illustrates much lower incident counts. These patterns suggest that factors such as accessibility, lack of surveillance, and the density of parked motorbikes significantly influence theft risks across different premises types. Incorporating this variable allows the model to account for environmental factors, improving its ability to predict theft occurrences based on location characteristics.

```{r}
#| label: plt-premises_summary
#| fig-cap: "Theft Incidents by Day of the Week"
#| echo: false
#| fig-width: 8
#| fig-height: 3.5
#| warning: false
#| message: false

@@

# Create the plot for premises type
ggplot(premises_summary, aes(x = reorder(premises_type, -theft_count), y = theft_count)) +
  geom_bar(stat = "identity", fill = "darkgrey") +
  theme_minimal() +
  labs(
    x = "Premises Type",
    y = "Number of Theft Incidents"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )

@@

The **Neighborhood ID** variable serves as a categorical predictor, uniquely identifying neighborhoods within the City of Toronto. This variable captures spatial differences in motorbike theft and collision risks, facilitating the identification of localized patterns and trends. @tbl-neighborhood_summary provides examples of Neighborhood IDs and their corresponding incident counts, illustrating the variation in theft incidents across different areas. This data highlights neighborhoods with high or low incident rates, enabling the model to incorporate spatial variability and improve predictions by understanding how collision and theft risks are distributed geographically.

@@

The **Road Surface Condition** variable serves as a categorical predictor, detailing the state of the road at the time of an incident. Categories include dry, wet, loose snow, slush, ice, packed snow, and others. This variable enables the analysis to assess how different surface conditions contribute to motorbike collision risks.

@plot-road_surface_summary illustrates the number of incidents under each road surface condition, showing that most incidents occur on dry roads, followed by wet roads. Conditions such as loose snow, ice, and packed snow account for a smaller proportion of incidents, reflecting their lower occurrence frequency. However, these different surface conditions may represent a disproportionately high risk per unit of exposure. This distribution underscores the importance of accounting for surface conditions in collision prevention strategies, especially during adverse weather, to enhance road safety and reduce risks.

```{r}
#| label: plot-road_surface_summary
#| fig-cap: "Incidents by Road Surface Condition"
#| echo: false
#| warning: false
#| message: false
#| fig-width: 8
#| fig-height: 3.5

@@

# Create the bar plot for all road surface conditions
ggplot(road_surface_summary, aes(x = reorder(road_conditions, -incident_count), y = incident_count)) +
  geom_bar(stat = "identity", fill = "darkgrey") +
  theme_minimal() +
  labs(
    x = "Road Surface Condition",
    y = "Number of Incidents"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )

@@

The **Lighting Condition** variable is a categorical predictor that describes the lighting environment at the time of an incident, including categories such as daylight, darkness, artificial lighting, and transitional periods like dawn and dusk. This variable helps analyze how visibility and lighting conditions impact the likelihood of motorbike thefts and collisions.

@plt-lighting_summary illustrates that the majority of incidents occur during daylight hours, likely due to higher traffic volumes and activity levels during the day. Incidents under "Dark" and "Dark, artificial" conditions are less frequent but still significant, highlighting the increased risks associated with poor visibility. Transitional conditions like "Dawn" and "Dusk" contribute minimally to the total number of incidents. These findings emphasize the need for targeted safety measures, such as enhancing artificial lighting and promoting driver vigilance in low-visibility conditions, to reduce collision risks effectively.

```{r}
#| label: plt-lighting_summary
#| fig-cap: "Incidents by Lighting Conditionk"
#| echo: false
#| warning: false
#| message: false
#| fig-width: 8
#| fig-height: 3.5
#| 
# Combine "NA" with "Dawn, artificial" and summarize
lighting_summary <- cleaned_traffic_data %>%
  mutate(
    lighting_conditions = case_when(
      lighting_conditions %in% c("NA", "Dawn, artificial") ~ "Dawn, artificial",
      TRUE ~ lighting_conditions
    )
  ) %>%
  filter(lighting_conditions != "Other") %>%  # Exclude "Other"
  group_by(lighting_conditions) %>%
  summarize(incident_count = n()) %>%
  arrange(desc(incident_count))  # Sort by the number of incidents

# Create the updated bar plot
ggplot(lighting_summary, aes(x = reorder(lighting_conditions, -incident_count), y = incident_count)) +
  geom_bar(stat = "identity", fill = "darkgrey") +
  theme_minimal() +
  labs(
    x = "Lighting Condition",
    y = "Number of Incidents"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )

@@

The **Traffic Control** variable categorizes the type of traffic management present at the site of each incident, such as "No Control," "Traffic Signal," or "Stop Sign." This variable provides insight into how various traffic control measures influence the likelihood of motorbike collisions.

As shown in @tbl-traffic_control_summary, the majority of incidents occur in areas with "No Control" (9,021 incidents), indicating that the absence of traffic regulation significantly contributes to collision risks. Areas with "Traffic Signal" also account for a substantial number of incidents (8,035), likely due to the higher traffic volumes typically associated with signalized intersections. Other traffic control types, such as "Stop Signs" (1,464 incidents) and "Pedestrian Crossovers" (208 incidents), are associated with considerably fewer incidents, reflecting their more localized application.

```{r}
#| label: tbl-traffic_control_summary
#| fig-cap: "Summary of Traffic Control"
#| echo: false
#| warning: false
#| message: false

@@

The **Time of Day** variable records the exact hour and minute when a collision occurred, represented in a 24-hour format. For example, "2216" corresponds to 10:16 PM, and "807" indicates 8:07 AM. This variable enables the analysis of temporal patterns in collision risks, identifying times of higher or lower incident probabilities.

Factors influencing these patterns may include reduced visibility during nighttime hours, increased traffic during peak commuting times, or variations in driver behavior throughout the day. @tbl-time_of_day_summary illustrates examples of incident counts at specific times, providing insights into how collisions are distributed across the day. Analyzing these temporal trends helps to identify high-risk periods, which can guide the implementation of targeted safety measures, such as increased monitoring or public awareness campaigns during specific hours.

```{r}
#| label: tbl-time_of_day_summary
#| fig-cap: "5 Examples of Different Times of Day with Incident Count"
#| echo: false
#| warning: false
#| message: false

# Load necessary libraries
library(tidyverse)
library(kableExtra)
library(here)

@@

# Select 5 random examples of time_of_day with their incident counts
time_of_day_summary <- cleaned_traffic_data %>%
  group_by(time_of_day) %>%
  summarize(incident_count = n()) %>%
  arrange(desc(incident_count)) %>%
  distinct() %>%               # Ensure unique times
  sample_n(5)                  # Randomly select 5 examples

# Create the table with a frame
time_of_day_summary %>%
  kable(
    col.names = c("Time of Day", "Incident Count"),
    align = "cc"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "bordered"),
    full_width = FALSE
  )
```

#### Visibility

The **Visibility** variable is a categorical predictor that captures environmental visibility conditions at the time of a collision, such as clear, rain, snow, fog, or strong winds. This variable is essential for understanding how varying visibility levels influence collision risks. Reduced visibility conditions, such as those caused by rain or snow, can impair drivers' ability to detect road hazards or other vehicles, increasing the likelihood of incidents. Conversely, clear conditions often correlate with safer driving environments.

@plot-visibility_summary summarizes the distribution of collisions across different visibility conditions, revealing that the vast majority of incidents occur during clear conditions. This is likely due to the overall higher frequency of clear weather compared to adverse conditions. Rain accounts for the second-highest number of incidents, while other conditions, such as snow, fog, and strong winds, contribute to a much smaller share of collisions.

```{r}
#| label: plot-visibility_summary
#| fig-cap: "Incidents by Visibility Condition"
#| echo: false
#| warning: false
#| message: false
#| fig-width: 8
#| fig-height: 3.5

@@

# Create the updated bar plot for visibility conditions
ggplot(visibility_summary, aes(x = reorder(visibility_conditions, -incident_count), y = incident_count)) +
  geom_bar(stat = "identity", fill = "darkgrey") +
  theme_minimal() +
  labs(
    x = "Visibility Condition",
    y = "Number of Incidents"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )

@@

```{r}
#| label: fig-risk-index-timeseries-year
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "Risk Index Variations in 2020 during Covid-19"
#| fig-pos: H
#| fig-width: 6
#| fig-height: 3.5

# Load necessary libraries
library(readr)
library(dplyr)
library(ggplot2)
library(lubridate)
library(here)

# Load the primary traffic data
cleaned_traffic_data <- read_csv(here("data", "02-analysis_data", "cleaned_traffic_data.csv"))

# Load the risk index data
risk_index_data <- read_csv(here("data", "02-analysis_data", "final_risk_index.csv"))

# Merge datasets using the common key "id"
merged_data <- cleaned_traffic_data %>%
  left_join(risk_index_data, by = "id")

# Convert `report_date` to Date format
merged_data <- merged_data %>%
  mutate(report_date = as.Date(report_date, format = "%Y-%m-%d"))

# Handle missing values in `risk_index` (choose one method):
# 1. Exclude rows with NA in `risk_index`
merged_data <- merged_data %>%
  filter(!is.na(risk_index))

# OR
# 2. Replace NA with the mean of `risk_index`
# merged_data <- merged_data %>%
#   mutate(risk_index = ifelse(is.na(risk_index), mean(risk_index, na.rm = TRUE), risk_index))

# Filter data for the year 2020 and calculate the average risk index per date
risk_trends_2020 <- merged_data %>%
  filter(year(report_date) == 2020) %>%
  group_by(report_date) %>%
  summarize(avg_risk_index = mean(risk_index, na.rm = TRUE))

# Add a month column for faceting
risk_trends_2020 <- risk_trends_2020 %>%
  mutate(month = month(report_date, label = TRUE))

# Create the plot
ggplot(risk_trends_2020, aes(x = report_date, y = avg_risk_index)) +
  geom_line(color = "#66471d", size = 0.8) +
  facet_wrap(~month, scales = "free_x") +  # Facet by month
  labs(
    x = "Month",
    y = "Average Risk Index",
    title = "Risk Index Variations in 2020 during Covid-19"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

@@

To better understand how environmental factors influence the Risk Index, we examine the distribution of the Risk Index under various road surface conditions. This analysis explores whether different surface types—such as dry, wet, icy, and others—affect the level of risk associated with traffic incidents. The following violin plot visually compares the Risk Index across different road surface conditions, providing insight into how these factors may contribute to the overall risk.
```{r}
#| label: fig-risk-distribution-roadcondition
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "Risk Index Distribution by Road Surface Conditions"
#| fig-pos: H
#| fig-width: 8.5
#| fig-height: 4

# Load necessary libraries
library(ggplot2)
library(dplyr)
library(readr)
library(here)

# Load data
cleaned_traffic_data <- read_csv(here("data", "02-analysis_data", "cleaned_traffic_data.csv"))
risk_index_data <- read_csv(here("data", "02-analysis_data", "final_risk_index.csv"))

# Merge datasets using the "id" column
merged_data <- cleaned_traffic_data %>%
  left_join(risk_index_data, by = "id")

# Combine "Spilled Liquid" with "Wet" and filter the data
violin_data <- merged_data %>%
  mutate(
    road_conditions = case_when(
      road_conditions == "Spilled liquid" ~ "Wet",
      TRUE ~ road_conditions
    )
  ) %>%
  filter(!is.na(road_conditions) & !is.na(risk_index))  # Ensure no NA values

# Define a custom color palette for the road conditions
custom_colors <- c(
  "Dry" = "#e8eef1",
  "Wet" = "#bfdff1",
  "Ice" = "#abd7f1",
  "Other" = "#96cff1",
  "Loose Snow" = "#6dc0f1",
  "Loose Sand or Gravel" = "#43b0f1",
  "Packed Snow" = "#057dcd",
  "Slush" = "#1e3d58"  
)

# Create the updated violin plot with transparency
ggplot(violin_data, aes(x = road_conditions, y = risk_index, fill = road_conditions)) +
  geom_violin(trim = FALSE, alpha = 0.7) +  # Add transparency with alpha
  scale_fill_manual(values = custom_colors) +  # Use custom colors
  theme_minimal() +
  labs(
    x = "Road Surface Condition",
    y = "Risk Index",
    fill = "Road Conditions"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )

@@

```{r}
#| label: fig-risk-index-lighting-conditions
#| fig-cap: "Risk Index by Lighting Conditions"
#| fig-subcap: ["Low Light (Dark)", "Dawn Light", "Bright Daylight and Artificial Light", "Artificial Lighting Conditions"]
#| echo: false
#| warning: false
#| message: false
#| fig-width: 8.5
#| fig-height: 3.5
#| layout-ncol: 2
#| layout-nrow: 2

@@

# First Plot: Dark
plot_dark <- ggplot(merged_data %>% filter(lighting_condition_group == "Dark"), aes(x = risk_index, fill = lighting_condition_group)) +
  geom_density(alpha = 0.6) +
  scale_fill_manual(values = c("Dark" = "#54350a")) +
  labs(x = "Risk Index", y = "Density") +
  theme_minimal() +
  theme(legend.position = "none")

# Second Plot: Daylight
plot_daylight <- ggplot(merged_data %>% filter(lighting_condition_group == "Daylight"), aes(x = risk_index, fill = lighting_condition_group)) +
  geom_density(alpha = 0.6) +
  scale_fill_manual(values = c("Daylight" = "#54350a")) +
  labs(x = "Risk Index", y = "Density") +
  theme_minimal() +
  theme(legend.position = "none")

# Third Plot: Dawn
plot_dawn <- ggplot(merged_data %>% filter(lighting_condition_group == "Dawn"), aes(x = risk_index, fill = lighting_condition_group)) +
  geom_density(alpha = 0.6) +
  scale_fill_manual(values = c("Dawn" = "#54350a")) +
  labs(x = "Risk Index", y = "Density") +
  theme_minimal() +
  theme(legend.position = "none")

# Fourth Plot: Artificial Lighting
plot_artificial <- ggplot(merged_data %>% filter(lighting_condition_group == "Artificial Lighting"), aes(x = risk_index, fill = lighting_condition_group)) +
  geom_density(alpha = 0.6) +
  scale_fill_manual(values = c("Artificial Lighting" = "#54350a")) +
  labs(x = "Risk Index", y = "Density") +
  theme_minimal() +
  theme(legend.position = "none")

@@

@fig-risk-index-lighting-conditions presents the distribution of the Risk Index across four distinct lighting categories: Dark Conditions, Daylight Conditions, Dawn Conditions, and Artificial Lighting Conditions.

Dark Conditions in (@fig-risk-index-lighting-conditions-1) highlights a concentration of Risk Index values in the moderate-to-high range, indicating that low visibility during dark conditions significantly contributes to elevated collision risks. These findings underscore the importance of improved street lighting or reflective road markers in mitigating risks under dark conditions.
